Demonstration of the solution of the system of mass and energy balances for MS Birka Main Engines

### 2017-11-26 Manuscript/Other/EngineBalance.tex
Added output for Sankey and Grassmann diagrams

%% 
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01

\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}
%% Tthe amsmath package provides the "align" environment for nicer equation formatting
\usepackage{amsmath}
%% The package tabls helps in having nicer tables with higher spaces
\usepackage{tabls}
%% The package multirow allows merging together multiple cells vertifcally
\usepackage{multirow}
%% 
\usepackage{subcaption}

\def\bibsection{\section*{References}}
%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\journal{Energy}

@@

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Energy and exergy analysis of a cruise ship}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author[EPFL]{Francesco Baldi}
\author[Linnaeus]{Fredrik Ahlgren}
\author[DTU]{Tuong-Van Nguyen}
\author[CTU]{Cecilia Gabrielii}
\author[CTU]{Karin Andersson}

\address[EPFL]{Industrial Process Energy Systems Engineering (IPESE), \'{E}cole Polytechnique F\'{e}d\'{e}rale de Lausanne, 1950, Sion, Switzerland}
\address[Linnaeus]{Kalmar Maritime Academy, Linnaeus University, Kalmar, Sweden}
\address[DTU]{Department of Mechanical Engineering, Technical University of Denmark, Lyngby, Denmark}
\address[CTU]{Department of Mechanics and Maritime Sciences, Chalmers University of technology, Gothenburg, Sweden}

\begin{abstract}
%% Text of abstract

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

low carbon shipping \sep energy analysis \sep exergy analysis \sep energy efficiency

\end{keyword}

\end{frontmatter}

%% \linenumbers

%% main text

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%%  \bibliographystyle{elsarticle-num} 
%%  \bibliography{<your bibdatabase>}

%% else use the following coding to input the bibitems directly in the
%% TeX file.

\section{Introduction} \label{sec:introduction}

\subsection{Background}

CO$_2$ emissions from shipping in 2012 amounted to a total of 949 million tonnes, contributing to 2.7\% of global anthropogenic CO$_2$ emissions \cite{Smith2014}. Although this contribution appears relatively low, the trend is that shipping will play an even greater role in the future due to the increased transport demand according to all IMO future scenarios \cite{Smith2014}. As an example, global transport demand has increased by 3.4\% in 2014, compared to a global GBP growth of 2.5\% the same year, which shows how shipping tends to rise even faster than global economy \cite{UNCTAD2015}. The OECD countries have reduced the CO$_2$ impact from shipping, but a larger amount has been moved to the non-OECD countries [3]. The fact that shipping needs to even further reduce its CO$_2$ emissions in the near future is essential for being able to achieve the goals of maintaining the climate below a 2-degree level in 2050 [4].

In addition to considerations related to GHG emissions, an emission control area (ECA) is enforced in the Baltic Sea  by the International Maritime Organisation since January 2015. This ECA stipulates that the fuel used must not contain more than 0.1\% sulphur, therefore requiring the use of more expensive distillate fuels. More generally making shipping sustainable is a challenge that will demand growing attention by the shipping industry \cite{Andersson2016}

Altogether, these conditions present a challenge to the shipping companies who attempt to reduce their fuel consumption, environmental impact, and operative costs. A wide range of fuel saving solutions for shipping are available and partially implemented in the existing fleet, both from the design and operational perspective; several specific studies have been conducted on these technologies, and a more detailed treatise would be out of the scope of this work. In this context, it has been acknowledged that the world fleet is heterogeneous, and measures need to be evaluated on a ship-to-ship basis \cite{Bouman2017}. In this process, a deeper understanding of energy use on board of the specific ship is vital.

\subsection{Previous work}

The idea of improving the understanding of the behavior of the energy system of a ship is not new. Most of the work published around this subject relates to the use  of mathematical models of the ship systems.

Most authors focused on the propulsion part of the problem, as this is often the most relevant energy demand on board. Shi et al. proposed a modeling approach for predicting ship fuel consumption of a cargo/passenger ferry \cite{Shi2009}; Theotokoats and Tzelepis applied a similar procedure to the case of a Handymax product carrier \cite{Theotokatos2015}, similarly to Tillig et al., who also added the dynamic element to their predictive model \cite{Tillig2016}.

The work referenced above allowed improving the understanding of how many operational (such as the ship's speed) and environmental (such as wave height and wind speed) parameters influence the ship's energy performance. These studies, however, are not based on actual measurements of the ship's operations. Also, they focus entirely on the ship's power demand for propulsion. This is a very reasonable practice for most ship types, given that propulsion represent the largest part of the total energy demand, but does not allow improving the knowledge of the remaining part of the system.

% Data-driven analysis
Other authors filled the gap by both including electric energy demand, and by basing their work on measurements from actual ship operations. The work presented by Thomas et al. \cite{Thomas2010} and Basurko et al. \cite{Basurko2013} shows the application of energy auditing methods to fishing vessels. This approach represents a step forward towards improving the understanding of how energy is used on board during actual ship operations.

More generally, several authors have highlighted the importance of a detailed knowledge of the ship's operational profile in order to appropriately assess and optimize possible alternatives for improving ship energy efficiency \cite{Banks2013}. Coraddu et al., started from a statistical analysis of measured ship operations and used the aggregated data, together with a computational model of the vessel, to provide a better prediction of the actual ship's operational efficiency \cite{Coraddu2014}. The importance of considering the operational profile was also showed in the case of the optimization of engine-propeller interaction \cite{Baldi2015c}, in the process of retrofitting existing systems \cite{Baldi2015b,Choi2013}, and in ship design \cite{Ghassemi2017,Solem2015}.

% Heat availability
Heat demand is rarely a subject of concern on board ships, with some notable exceptions. This is due to a combination of generally low demand and high availability from the waste heat of the engines. In a previous publications by the authors, for instance, it is shown that in the case of a product tanker, although the heat demand was estimated to account for roughly 20\% of the total energy demand of the ship on a yearly basis, it only contributes to 4.1\% of the total fuel consumption (contribution of the auxiliary boilers), while the rest of the demand is fulfilled using waste heat \cite{Baldi2015a}.

%% WHR and heat availability
It should be noted that, however, much research effort has been devoted, especially in recent times, to the improvement of the efficiency of ship energy systems by recovering the waste heat available from the engines. With reference to different types of technologies, case studies, and designs,the several authors showed the existance of a quite significant potential for energy saving when WHR systems are employed, ranging from around 1\% for single-pressure steam cycles applied to two-stroke engines \cite{Theotokatos2012} to more complex systems based on ORCs (up to 10\%, \cite{Hountalas2012}) or including the cooling systems as a source of waste heat (over10\%, \cite{Dimopoulos2012}).  The case of the installation of an ORC on board of the vessel investigated in this study (see Section \ref{sec:met:case}) was presented in \cite{Mondejar2015}, showing a similar potential.

The potential uses for waste heat on board are not limited to improving the efficiency of the power plant. Waste heat is commonly used for fulfilling on board heat demand for spatial heating and freshwater generation \cite{Molland2011,Baldi2015a,Mccarthy1990};  Balaji et al. proposed the use of waste heat for ballast-water systems \cite{Balaji2012}; Salmi et al. suggested its use for adsorption refrigeration systems \cite{Salmi2017}. A detailed review of potential uses for waste heat from marine engines is presented by Shu et al \cite{Shu2013}.

% Cruise ships
Some ship types constitute notable examples. On cruise ships the heat demand is significantly higher compared to standard cargo ships. Referring to winter conditions, Marty et al. estimated the instantaneous heat demand of a selected cruise ship to reach roughly 23 MW, compared to an estimated peak of 49 MW for propulsion and electric auxiliaries combined \cite{Marty2012}.

This shows how the heat from the engines should be considered as a potential resource, rather than a waste, and that there is potential for improvement based on the optimal use of these heat sources. The work presented by the authors in \cite{Baldi2016} represent a step in this direction; however, there is the need for an increased detail in the estimation of the heat demand.

% Exergy analysis
Exergy analysis provides a more accurate estimation of the potential for energy recovery on board. The application of exergy analysis to the case of ship energy systems is, however, still limited. Dimopoulos et al. showed how the process optimizing the WHR system of a container ship can be more efficient if exergy efficiency, rather than energy efficiency, is used as the target of the optimization \cite{Dimopoulos2012}; Baldi et al. also analyzed the exergy flows on board of a product tanker, showing that this allows having a more accurate understanding of what parts of the system show potential for improvement \cite{Baldi2015a}; similar results were obtained by Marty et al., who focused on the power plant of a cruise ship \cite{Marty2016}. Koroglu et al. made a step futher also including advanced exergy analysis in their study \cite{Koroglu2017}.

\subsection{Aim}

Given the existing limitations of the current available information in scientific literature highlighted in the previous section, the objective of this work is the following:
\begin{itemize}
	\item Analyse the demand of a cruise ship in terms of propulsion power, electric power and heat, based on operational measurements. The lack of existing information with reference to the heat demand is considered a particular element of novelty when compared to existing literature in the subject.
	\item Analyse the current efficiency of the system, and potential ways to improve it, by means of applying exergy analysis.
	\item Provide reference operational conditions for further use in the field of energy systems optimization.
\end{itemize}

\section{Method} \label{sec:method}

In this paper, we present the application of energy and exergy analysis (described in section \ref{sec:met:energyExergy}) to a cruise ship. This is done for one specific case study vessel, that is described in detail in Section \ref{sec:met:case}; in Section \ref{sec:met:gathering} we present the specifics of the available information, in particular the measured data and the technical documentation; the specific assumptions and methods used for processing the available information into  energy and exergy flows are specified in Section \ref{sec:met:processing}. In these regards, the estimation of the heat demand is treated in a separate section (Sec. \ref{sec:met:heat}) as it constitutes a particularly challenging task and it is one of the central aspects of this article.

\subsection{Energy and exergy analysis} \label{sec:met:energyExergy}

\subsubsection{Energy analysis}

\emph{Energy} can be stored, transformed from one form to another (e.g. heat to power) and transferred between systems, but can neither be created nor destroyed (conservation law)~\textbf{Bejan et al.}. The system under study is the \emph{ship energy system} and is thus taken as control volume. The energy balance can be expressed as:

\begin{align}
	\sum_{\mathrm{in}} \dot{H}_{\mathrm{in}} &= \sum_{\mathrm{out}} \dot{H}_{\mathrm{out}} \\
	\dot{H}_{\mathrm{fuel}} + \dot{H}_{\mathrm{air}} &= \sum_{{\mathrm{waste}}}\dot{H}_{{\mathrm{waste}}}+\dot{W}_{\mathrm{el}}+\dot{Q}_{\mathrm{heating}}
\end{align}

The left-hand side term represents, on a time rate basis, the energy associated with the fuel consumed in the boilers and engines ($\dot{H}_{\mathrm{fuel}}$) and the air used in the combustion processes ($\dot{H}_{\mathrm{air}}$). The right-hand side term denotes the power ($\dot{W}_{\mathrm{el}}$) and heat ($\dot{Q}_{\mathrm{heating}}$) required on-site (e.g. propulsion and fuel heating), and the heat discharged into the environment ($\sum_{{\mathrm{waste}}}\dot{H}_{{\mathrm{waste}}}$) with, for instance, the exhaust gases.

The energy flow associated with a material stream is calculated as the sum of the physical and chemical enthalpies, and kinetic and potential energies are neglected.    The physical energy is taken as the relative enthalpy, as underlined in~\textbf{Kotas et al.}, and the chemical energy is taken as the lower/higher heating value. The environmental conditions taken for the present analysis are the ambient pressure and seawater temperature.

\subsubsection{Exergy analysis}

\emph{Exergy} may be defined as the `maximum theoretical useful work (shaft work or electrical work) as the system is brought into complete thermodynamic equilibrium with the thermodynamic environment while the system interacts with it only'~\citep{Moran1989}. Unlike energy, exergy is not conserved but some is destroyed because of the irreversible phenomena taking place in real processes (e.g. chemical reactions like combustion). The exergy balance for the system under study can be expressed as:

\begin{align}
	\sum_{\mathrm{in}} \dot{E}_{\mathrm{in}} &= \sum_{\mathrm{out}} \dot{E}_{\mathrm{out}} +\dot{E}_{d} \\
	\dot{E}_{\mathrm{fuel}} + \dot{E}_{\mathrm{air}} &= \sum_{{\mathrm{waste}}}\dot{E}_{{\mathrm{waste}}}+\dot{E}_{W}+\dot{E}_{Q,\mathrm{heating}}+\dot{E}_{d} 
\end{align}

The left-hand side term represents, on a time rate basis, the exergy associated with the fuel and air. The right-hand side term denotes the exergy of the waste streams, the exergy transfers with heat and power, and the exergy destroyed in the ship system. Kinetic and potential exergies are neglected, and the exergy of a given material stream is derived as the sum of the physical and chemical exergies. The chemical exergy is calculated based on the reference environment of \textbf{Szargut (1989)}, and its value is approximatively equal to the higher heating value for hydrocarbon fuels. The exergy destruction can be calculated from the Gouy-Stodola theorem~\textbf:{Kotas (1995)}.

The exergy balance may alternatively be formulated as:
\begin{equation}
\dot{E}_{\mathrm{p}} = \dot{E}_{\mathrm{f}} - \dot{E}_{\mathrm{d}} - \dot{E}_{\mathrm{l}} 
\end{equation}

where $\dot{E}_{\mathrm{p}}$ is called the exergy product, and corresponds to the desired output of the system, in exergy terms (for example, the power produced in an engine). $\dot{E}_{\mathrm{f}}$ denotes the exergy fuel, and represents the resources spent to drive the studied process (for instance, the fuel used in a combustion process). The last term $\dot{E}_{\mathrm{l}}$ corresponds to the losses of a system, such as the heat discharged into the environment with cooling water.

\subsubsection{System performance}

The following indicators are used to evaluate the system performance:
\begin{itemize}
	\item the exergy efficiency $\varepsilon$, defined as the ratio between the exergy product and fuel of a given component or system
	\begin{equation} \varepsilon = \frac{\dot{E}_{\mathrm{p}}}{\dot{E}_{\mathrm{f}}} \end{equation}
	\item the efficiency defect $\lambda$, presented in the work of Kotas, defined as the fraction of the total exergy input destroyed in the successive irreversible processes
	\begin{equation} \lambda = \frac{\dot{E}_{\mathrm{d}}}{\dot{E}_{\mathrm{in,tot}}} \end{equation}
	\item  the irreversibility share $\delta$, suggested in the work of Tsatsaronis, defined as the ratio between the exergy destroyed in the $i$-th component in relation to the exergy destroyed in the entire system
	\begin{equation} \delta_i = \frac{\dot{E}_{\mathrm{d},i}}{\dot{E}_{\mathrm{d,tot}}} \end{equation}

\end{itemize}

\subsection{Case study vessel} \label{sec:met:case}

The energy and exergy analysis are here applied to a specific  cruise ship operating daily cruises in the Baltic Sea between Stockholm and the island of \.{A}land. The ship is 176.9 m long and has a beam of 28.6 m, and has a design speed of 21 knots. The ship was built in Aker Finnyards, Raumo Finland in 2004.

The ship has a capacity of 1800 passengers and has several restaurants, night clubs and bars, as well as saunas and pools. This means that the energy system regarding the heat and electricity demand is more complex than a regular cargo vessel in the same size. Typical ship operations, although they can vary slightly between different days, are represented in Figure \ref{fig:typicalShipOperations}. It should be noted that the ship stops and drifts in open sea during night hours before mooring at its destination in the morning, if allowed by weather conditions.

The ship systems are summarized in Figure \ref{fig:shipSystems}. The propulsion system is composed of two equal propulsion lines, each made of two engines, a gearbox, and a propeller. The main engines are four Wärtsilä 4-stroke Diesel engines (ME) rated 5850 kW each. All engines are equipped with selective catalytic reactors (SCR) for NOX emissions abatement. Propulsion power is needed whenever the ship is sailing; however, it should be noted that the ship rarely sails at full speed, and most of the time it only needs one or two engines operated simultaneously.

Auxiliary power is provided by four auxiliary engines (AE) rated 2760 kW each. Auxiliary power is needed on board for a number of alternative functions, from pumps in the engine room to lights, restaurants, ventilation and entertainment for the passengers.

Auxiliary heat needs are fulfilled by the exhaust gas steam generators (HRSG) located on all four AEs and on two of the four MEs or by oil-fired auxiliary boilers (mainly when in port, or during winter), by the heat recovery on the HT cooling water systems (HRHT), and by the auxiliary, oil fired boilers (AB). The heat is needed for passenger and crew accommodation, as well as for the heating of the highly viscous heavy fuel oil used for engines and boilers. This last part, however, is drastically reduced since the 1st of January 2015, as new regulations entering into force require the use of low-sulphur fuels, which require a much more limited heating.

\subsection{Data gathering and pre-processing} \label{sec:met:gathering}

The operational data was collected on board from the ships' machine logging and surveillance system. The on board database tool exported all logging points to Excel-97 files, and due to the extensive amount of data points the export was divided in to 15 individual Excel-files consisting of a total 665 MB. The exported raw-data from the ship was over a time span of a full year and in most data-points in 15-min average. The Excel-files were processed in the Pandas library which is a high performance data analysis tool in Python \cite{mckinney-proc-scipy-2010}. A new structured naming of headers and a consistent time frequency over all data points were created, and 245 selected data points were saved in a HDF5 table time series database which is the base for the analysis.

All data-points were checked individually by creating a descriptive statistic and a histogram of each data-point. A filter was applied by setting up a pre-defined maximum and minimum value. As we do not know, or had any meaningful way of deriving, each individual sensors placement, sensor type or calibration status all data must be filtered and checked accordingly for outliers or bad data.

Since measurements of ambient and seawater temperature from onboard logging systems were not available for the whole dataset, we used measurements taken from SMHI database for Landsort lighthouse for the seawater, and the lighthouse Svenska H\"{o}garna for the ambient temperature. The Landsord lighthouse is situated south of the Stockholm archipelago, and the Svenska H\"{o}garna lighthouse is along the ship route, in between the Swedish archipelago and Åland. The assumption was validated based on  June-December period, for which onboard measurements were available. This resulted in a root mean square error of 1.5 K and 1.9 K for the seawater and the ambient temperature respectively, which we considered to be accurate enough for the purpose of this work. The fit the SMHI-data with the rest of the data-set the SMHI data was resampled from 1h for the seawater and 3h for the ambient temperature, to 15min frequency using a linear interpolation.

\subsection{Data processing: Ship energy system modeling} \label{sec:met:processing}

Not all of the variables required to perform a full energy and exergy analysis of the system are available from measurements. In some cases, they are measurable, but not measured (e.g. some temperatures, mass flows, etc.). In other cases, they are simply impossible, or impractical, to measure (e.g. specific enthalpy, specific entropy). For this reason, part of the ship systems needed to be modeled in order to derive the unknown variables.

\subsubsection{Diesel engines}

The setup of the engines and their connections to the other ship systems are summarized in Figure \ref{fig:DieselEngines} for the main and auxiliary engines, respectively. The most relevant operative values are provided in Tables \ref{tab:ME_values} and \ref{tab:AE_values} and the main energy flows are listed in Tables \ref{tab:ME_flows} and \ref{tab:AE_flows}. It should be noted that, in the case of the main engines, the engine power output was not measured and needed to be estimated. In this work, we calculated the engine power output based on measurements of engine fuel rack position (used as a proxy of the mass of fuel injected per cycle) and of the engine speed:
\begin{eqnarray}
\dot{m}_{fuel} & = & \dot{m}_{fuel,des} \left(a_0 + a_1 \frac{frp}{frp_{des}}\right) \frac{\omega}{\omega_{des}} \\
\eta_{ME} & = & a_0 + a_1 \frac{\dot{m}_{fuel}}{\dot{m}_{fuel,des}} + a_2 \left( \frac{\dot{m}_{fuel}}{\dot{m}_{fuel,des}} \right)^2 \\
\dot{W}_{ME} & = & \dot{m}_{fuel} \eta_{ME} LHV
\end{eqnarray}

The validity of this assumption can be seen by observing Figures \ref{fig:Pme_vs_vship_Neng} and \ref{fig:Pme_vs_TCspeed} where the calculated  power from the main engines is represented against the ship speed and the turbocharger speed, respectively.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{Figures/DieselEngines}
	\caption{Schematic representation of the Diesel engines and their connections to the cooling systems}
	\label{fig:DieselEngines}
\end{figure}

In the case of the auxiliary engines, instead, direct measurements of the power generated by each engine were available.

For both auxiliary and main engines, part of the relevant internal variables are not measured (particularly with relation to the bypass valves) and this required to determine them based on modeling the heat and mass balance of the engine. The model is explained in detail in Appendix REF.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{Figures/Pme_vs_vship_Neng}
	\caption{Scatter plot, Propulsion power versus ship speed}
	\label{fig:Pme_vs_vship_Neng}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{Figures/Scatter_Pme_vs_TCspeed}
	\caption{Scatter plot, Main engine power versus turbocharger speed (ME1)}
	\label{fig:Pme_vs_TCspeed}
\end{figure}

\begin{table}
	\small
	\centering
	\begin{tabular}{llccrrrr}
		\hline
		\multicolumn{2}{l}{Variable} & Unit & M/C & 30\% & 50\% & 70\% & 90\% \\
		\hline
		\multicolumn{2}{l}{Compressor} & & & & & & \\ 
		& 	$T_{air,in}$  		& $^o$C & M & \multicolumn{4}{c}{302} \\ 
		&	$T_{air,out}$  		& $^o$C & C  & 356.0 & 409.4 & 434.1 & 461.7 \\ 
		&	$\dot{m}_{air}$  	& $kg/s$ & C  & 4.24 & 7.62 & 8.51 & 10.28 \\
		&   $\beta_{comp}$		& - 	& M & & & &  \\
		\multicolumn{2}{l}{Turbine} & & & & & &  \\
		&	$T_{eg,in}$   		& $^o$C & C  & 698.8 & 702.2 & 763.3 & 786.0 \\ 
		&	$T_{eg,out}$  		& $^o$C & M  & 639.3 & 592.9 & 633.2 & 631.0 \\ 
		&	$\dot{m}_{eg}$  	& $kg/s$ & C  & 4.35 & 7.79 & 8.73 & 10.57 \\
		&   $\beta_{exp}$		& - 	& C & & & &  \\
		\multicolumn{2}{l}{Bypass valve} & & & & & &  \\
		&	$\dot{m}_{air,BP}$	& $kg/s$ & C  & 0.98 & 2.05 & 1.02 & 0.39 \\  
		\multicolumn{2}{l}{Cylinders} & & & & & &   \\
		&	$\dot{m}_{air,in}$  & $kg/s$ & C  & 3.25 & 5.57 & 7.49 & 9.89 \\
		&	$\dot{m}_{fuel,in}$ & $kg/s$ & C  & 0.11 & 0.17 & 0.22 & 0.28 \\ 
		&	$\dot{m}_{eg,out}$  & $kg/s$ & C  & 3.37 & 5.74 & 7.71 & 10.17 \\
		&	$T_{air,in}$  		& $^o$C & M  & 323.7 & 323.2 & 323.1 & 323.5 \\ 
		&	$T_{eg,out}$  		& $^o$C & M  & 698.8 & 702.2 & 763.3 & 786.0 \\ 
		\multicolumn{2}{l}{CAC-LT stage} & & & & & &  \\
		&	$T_{air,in}$  		& $^o$C & C  & 355.2 & 350.1 & 360.0 & 368.7 \\ 
		&	$T_{air,out}$ 		& $^o$C & M  & 323.7 & 323.2 & 323.1 & 323.5 \\
		&	$T_{w,in}$   		& $^o$C & M  & 309.3 & 309.6 & 309.5 & 309.6 \\
		&	$T_{w,out}$   		& $^o$C & C  & 310.4 & 311.0 & 311.8 & 313.0 \\  
		& 	$\dot{m}_{w,LT}$  	& $kg/s$ & C& 23.36 & 26.09 & 29.1 & 31.53 \\
		\multicolumn{2}{l}{CAC-HT stage} & & & & & & \\
		&	$T_{air,in}$   		& $^o$C & C  & 356.0 & 409.4 & 434.1 & 461.7 \\
		&	$T_{air,out}$   	& $^o$C & C  & 355.2 & 350.1 & 360.0 & 368.7 \\ 
		&	$T_{w,in}$    		& $^o$C & C  & 360.8 & 360.2 & 359.6 & 358.7 \\ 
		&	$T_{w,out}$   		& $^o$C & C  & 360.8 & 362.6 & 363.6 & 365.4 \\
		& 	$\dot{m}_{w,HT}$  	& $kg/s$ & C & 33.21 & 33.18 & 33.33 & 33.33 \\
		\multicolumn{2}{l}{LOC} & & & & & &  \\
		&	$T_{lo,in}$   		& $^o$C & C  &346.6 & 348.9 & 349.8 & 350.5 \\ 
		&	$T_{lo,out}$    	& $^o$C & M  & 335.1 & 336.8 & 338.1 & 338.3 \\ 
		&	$T_{w,in}$    		& $^o$C & C  & 310.4 & 311.0 & 311.8 & 313.0 \\
		&	$T_{w,out}$   		& $^o$C & C  & 317.4 & 317.6 & 317.5 & 318.5 \\ 
		\multicolumn{2}{l}{JWC} & & & & & &  \\
		&	$T_{wall}$    		& $^o$C & C  & \multicolumn{4}{c}{423}  \\ 
		&	$T_{w,in}$    		& $^o$C & M  & 358.1 & 358.4 & 357.6 & 356.4 \\ 
		&	$T_{w,out}$	   		& $^o$C & C  & 360.8 & 360.2 & 359.6 & 358.7 \\ 
		\hline
		\end{tabular}
	\caption{Main engines, measured and calculated temperatures and flows at different engine loads}
	\label{tab:ME_values}
\end{table}

INCLUDE VALIDATION AGAINST PROJECT GUIDE DATA

\begin{table}
	\small
	\centering
	\begin{tabular}{llccrrrr}
		\hline
		\multicolumn{2}{l}{Variable} & Unit & M/C & 30\% & 50\% & 70\% & 90\% \\
		\hline
		\multicolumn{2}{l}{Compressor} & & & & & & \\ 
		& 	$T_{air,in}$  		& $^o$C & M & 284.2 & 285.3 & 284.9 & 294.6 \\ 
		&	$T_{air,out}$  		& $^o$C & C  & 323.2 & 357.9 & 391.6 & 430.9 \\
		&	$\dot{m}_{air}$  	& $kg/s$ & C  & 1.96 & 2.65 & 3.5 & 4.27 \\
		&   $\beta_{comp}$		& - 	& M & & & &  \\
		\multicolumn{2}{l}{Turbine} & & & & & &  \\
		&	$T_{eg,in}$   		& $^o$C & M  & 729.5 & 762.1 & 770.3 & 798.5 \\
		&	$T_{eg,out}$  		& $^o$C & M  & 652.8 & 667.8 & 654.2 & 659.4 \\
		&	$\dot{m}_{eg}$  	& $kg/s$ & C  & 2.02 & 2.74 & 3.61 & 4.41 \\
		&   $\beta_{exp}$		& - 	& C & & & &  \\
		\multicolumn{2}{l}{Cylinders} & & & & & &   \\
		&	$\dot{m}_{air,in}$  & $kg/s$ & C  & 1.96 & 2.65 & 3.5 & 4.27 \\ 
		&	$\dot{m}_{fuel,in}$ & $kg/s$ & C  & 0.05 & 0.08 & 0.11 & 0.14 \\
		&	$\dot{m}_{eg,out}$  & $kg/s$ & C  & 2.02 & 2.74 & 3.61 & 4.41 \\
		&	$T_{air,in}$  		& $^o$C & M  & 320.8 & 322.3 & 327.6 & 329.1 \\
		&	$T_{eg,out}$  		& $^o$C & C  & 729.5 & 762.1 & 770.3 & 798.5 \\
		\multicolumn{2}{l}{CAC-LT stage} & & & & & &  \\
		&	$T_{air,in}$  		& $^o$C & C  & 323.2 & 356.3 & 366.7 & 372.0 \\ 
		&	$T_{air,out}$ 		& $^o$C & M  & 320.8 & 322.3 & 327.6 & 329.1 \\ 
		&	$T_{w,in}$   		& $^o$C & M  & 317.1 & 316.3 & 317.5 & 314.1 \\
		&	$T_{w,out}$   		& $^o$C & C  & 317.2 & 318.2 & 320.5 & 318.3 \\
		& 	$\dot{m}_{w,LT}$  	& $kg/s$ & C & 11.16 & 11.16 & 10.98 & 10.78 \\ 
		\multicolumn{2}{l}{CAC-HT stage} & & & & & & \\
		&	$T_{air,in}$   		& $^o$C & C  & 323.2 & 357.9 & 391.6 & 430.9 \\ 
		&	$T_{air,out}$   	& $^o$C & C  & 323.2 & 356.3 & 366.7 & 372.0 \\ 
		&	$T_{w,in}$    		& $^o$C & C  & 362.5 & 364.2 & 368.0 & 368.2 \\
		&	$T_{w,out}$   		& $^o$C & C  & 362.5 & 364.3 & 369.3 & 371.9 \\
		& 	$\dot{m}_{w,HT}$  	& $kg/s$ & C & 16.66 & 16.67 & 16.55 & 16.67 \\
		\multicolumn{2}{l}{LOC} & & & & & &  \\
		&	$T_{lo,in}$   		& $^o$C & C  & 348.2 & 350.3 & 354.0 & 357.3 \\
		&	$T_{lo,out}$    	& $^o$C & M  & 336.6 & 337.4 & 339.7 & 340.4 \\
		&	$T_{w,in}$    		& $^o$C & C & 317.2 & 318.2 & 320.5 & 318.3 \\ 
		&	$T_{w,out}$   		& $^o$C & C  & 325.9 & 327.8 & 331.4 & 331.3 \\
		\multicolumn{2}{l}{JWC} & & & & & & \\
		&	$T_{wall}$    		& $^o$C & C  & \multicolumn{4}{c}{423}  \\
		&	$T_{w,in}$    		& $^o$C & M  & 360.0 & 359.7 & 362.3 & 361.5 \\
		&	$T_{w,out}$	   		& $^o$C & C  & 362.5 & 364.2 & 368.0 & 368.2 \\  
		\hline
	\end{tabular}
	\caption{Auxiliary engines, measured and calculated temperatures and flows at different engine loads}
	\label{tab:AE_values}
\end{table}

INCLUDE VALIDATION AGAINST PROJECT GUIDE DATA

\begin{table}
	\small
	\centering
	\begin{tabular}{lcrrrr}
		\hline
		Energy flow & type & 30\% & 50\% & 70\% & 90\% \\
		\hline
		Power output & Mech &  1767 & 2946 & 4118 & 5293 \\	
		Exhaust gas (after turbine) & Heat & 1672 & 2566 & 3227 & 3938 \\
		CAC-LT & Heat & 104 & 152 & 279 & 454 \\
		CAC-HT & Heat & 2 & 337  & 569 & 843  \\
		JWC & Heat & 381 & 249 & 278 & 318 \\
		LOC & Heat & 682 & 721 & 699 & 725  \\
		\hline
	\end{tabular}
	\caption{Main engines, energy flows at different engine loads}
	\label{tab:ME_flows}
\end{table}

\begin{table}
	\small
	\centering
	\begin{tabular}{lcrrrr}
		\hline
		Energy flow & type & 30\% & 50\% & 70\% & 90\% \\
		\hline
		Power output & Mech & 828 & 1381 & 1932 & 2482 \\

		Exhaust gas (after turbine) & Heat & 808 & 1144 & 1456 & 1753 \\
		CAC-LT & Heat & 5 & 91 & 139 & 216 \\
		CAC-HT & Heat & 0 & 5 & 89 & 257  \\
		JWC & Heat & 176 & 314 & 396 & 467 \\ 
		LOC & Heat & 405 & 448 & 499 & 587  \\
		\hline
	\end{tabular}
	\caption{Auxiliary engines, energy flows at different engine loads}
	\label{tab:AE_flows}
\end{table}

\subsubsection{Cooling systems}

The ship's cooling systems are designed similarly to most ships and divided in two cooling systems of two separate temperature levels: the high temperature (HT) cooling systems (70-90$^o$C) and the low temperature (LT) cooling systems (40-60$^o$C). The cooling systems provide the necessary cooling to the engines, to the steam systems and to all other components on board. A representation of the ship cooling systems is provided in Figure XXX.

All values not directly measured in the cooling systems were calculated based on mass and energy balances.

Figure cooling systems

\subsection{Estimation of the electric power demand} \label{sec:met:electric}

>>>>

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{verbatim}

%opening
\title{Demonstration of the solution of the system of mass and energy balances for MS Birka Main Engines}
\author{Francesco Baldi}

@@

\maketitle

\begin{abstract}

@@

\section{Initial system}
